---
title: "Connecting MERMAID ecological data with covariates example"
subtitle: "Exploring relationships between hard coral cover and sea surface temperature across marine realms"
author:
  - name: Iain R. Caldwell
  - name: Sharla Gelfand
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    self-contained: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
title-block-banner: "#f0f3f5"
title-block-banner-color: "black"
include-after-body: "footer.html"
---

## Overview

This analysis explores the relationship between hard coral cover and sea surface temperature (SST) across different marine realms using global MERMAID benthic data. We use the [mermaidrcovariates](https://data-mermaid.github.io/mermaidr-covariates/) package to access environmental data for MERMAID survey locations.

**Key questions:**

- How does hard coral cover vary with maximum SST?
- Do these relationships differ across marine biogeographic realms?

## Setup

```{r setup}
#| warning: false
library(here)
library(tidyverse)
library(mermaidr)
library(plotly)
library(DT)

# Install mermaidr.covariates (if needed)
remotes::install_github("data-mermaid/mermaidr-covariates")

library(mermaidrcovariates)

# Set theme for ggplot
theme_set(theme_minimal(base_size = 12))

# Set this to TRUE to re-download everything instead of using the cached files
force_download <- FALSE
```

## Data Export

### MERMAID Benthic Data

Export all publicly available summary sample events from MERMAID.

```{r get-mermaid-data}
#| results: asis
# Get all public benthic data at the sample event level (all protocols).
# Cached to data/ on first render. This is the same global export the DHW
# example saves before filtering it, so the two documents share one file.
mermaid_summ_ses_path <- here::here("data", "mermaid_summ_ses.rds")

if (file.exists(mermaid_summ_ses_path) && !force_download) {
  mermaid_summ_ses <- readRDS(mermaid_summ_ses_path)
  cat('<span style="color: #4682B4;">Loaded cached sample events from data/mermaid_summ_ses.rds</span>\n\n')
} else {
  mermaid_summ_ses <- mermaid_get_summary_sampleevents()
  dir.create(dirname(mermaid_summ_ses_path), showWarnings = FALSE, recursive = TRUE)
  saveRDS(mermaid_summ_ses, mermaid_summ_ses_path)
}

# Check how many sample events we have
cat('<span style="color: #006400;">Total sample events:', nrow(mermaid_summ_ses), '</span>\n\n')
```

### Extract Hard Coral Cover

Hard coral cover is stored in the percent cover columns. We'll focus on total hard coral cover and show a histogram of the data.

```{r extract-coral-cover}
# Extract relevant columns
coral_data <- mermaid_summ_ses %>% 
  rowwise() %>% 
  mutate(hard_coral_cover =
           mean(c(`benthicpit_percent_cover_benthic_category_avg_Hard coral`,
                  `benthiclit_percent_cover_benthic_category_avg_Hard coral`,
                  `benthicpqt_percent_cover_benthic_category_avg_Hard coral`,
                  quadrat_benthic_percent_percent_hard_avg_avg), na.rm = T)) %>% 
  ungroup() %>% 
  filter(!is.na(hard_coral_cover)) %>% 
  select(
    project,
    country,
    site,
    latitude,
    longitude,
    sample_date,
    hard_coral_cover
  ) %>% 
  # Convert date to Date format
  mutate(sample_date = as.Date(sample_date))

# Check how many sample events have hard coral cover estimates 
cat('<span style="color: #006400;">Total sample events with hard coral data:', nrow(coral_data), '</span>\n\n')

# Create the histogram to get bin data to assign colors
hist_data <- hist(coral_data$hard_coral_cover,
                  breaks = seq(0, 100, by = 2),
                  plot = FALSE) #In this function a 4 will be in the first bin

# Assign colors based on bin order
bin_colors <- sapply(seq_along(hist_data$counts), function(i) {
  if (i <= 5) {
    return('#d13823')
  } else if (i > 5 && i <= 15) {
    return('#f3a224')
  } else {
    return('#277d1d')
  }
})

# Create the histogram using plotly
hardCoralAggHist <-
  plot_ly(x = coral_data$hard_coral_cover,
          type = 'histogram',
          xbins = list(start = 0, size = 2, end = 100),
          marker = list(color = bin_colors), height = 450,
          hovertemplate = "Bin: %{x}%<br>%{y} surveys<extra></extra>") %>%
  config(displayModeBar = TRUE,
         displaylogo = FALSE,
         modeBarButtonsToRemove = c('zoom','pan', 'select', 'zoomIn', 'zoomOut',
                                    'autoScale', 'resetScale', 'lasso2d',
                                    'hoverClosestCartesian',
                                    'hoverCompareCartesian')) %>% 
  layout(bargap = 0.1,
         shapes = list(
           list(type = "line", 
                x0 = 10, x1 = 10, y0 = 0, y1 = 1, yref = "paper", 
                line = list(color = "black", dash = "dot")),
           list(type = "line", 
                x0 = 30, x1 = 30, y0 = 0, y1 = 1, yref = "paper", 
                line = list(color = "black", dash = "dot"))
         ),
         xaxis = list(title = "Hard coral cover (%)",
                      linecolor = "black",
                      linewidth = 2,
                      tickvals = seq(0, 100, by = 10),  # Set x-axis tick values
                      ticktext = seq(0, 100, by = 10)),
         yaxis = list(title = "Number of surveys",
                      linecolor = "black",   # Set the y-axis line color to black
                      linewidth = 2),
         annotations = list(
           list(x = 0, y = 1.15, text = "HARD CORAL COVER", showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 20)),
           list(x = 0, y = 1.08,
                text = paste0(nrow(coral_data), " Surveys"),
                showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 12))
         ),
         margin = list(t = 50, b = 75)) # Increase top margin to create more space for title and subtitle

# Visualize the plot
hardCoralAggHist
```

### Get Covariates

Now we'll use `mermaidrcovariates` to get SST and Marine Ecoregions of the World (MEOW) realm data.

#### List covariates

The first function of interest is [`list_covariates`](https://data-mermaid.github.io/mermaidr-covariates/reference/list_covariates.html), which can be used to find out which covariates are available.

```{r list-covariates}
# Extract id and title from list_covariates() output
covariates <- list_covariates()

covariates_df <- covariates %>% 
  select(id, title, description)

# Create interactive table
datatable(
  covariates_df,
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  caption = "Available MERMAID Covariates",
  rownames = FALSE,
  filter = "top"  # Adds search boxes at the top of each column
)
```

#### Extract and spatial join MEOW boundaries (vector)

Vector covariates such as MEOW are attached with [`attach_covariate_data()`](https://data-mermaid.github.io/mermaidr-covariates/reference/attach_covariate_data.html), which matches each sample event to the polygon containing it in a single call. This is a location-based join, so no date window is involved. Use [`list_datasets_for_covariate()`](https://data-mermaid.github.io/mermaidr-covariates/reference/list_datasets_for_covariate.html) to see which columns a vector covariate offers.

Note that this reads the covariate through DuckDB and needs the `duckdb` package at version 1.5 or later, since the covariate parquet uses native spatial geometry with CRS type modifiers. On Windows that in turn needs R 4.5 or newer, because CRAN only builds binaries for the current and previous R releases.

```{r extract-join-vector}
#| results: asis
# MEOW column names are upper case; rename to match the rest of this document.
coral_meow <- coral_data %>%
  attach_covariate_data(
    covariate = "meow_boundaries",
    columns = "REALM") %>%
  rename(realm = REALM)

# Determine if any of the sample events do not have a realm assigned and, if so, remove.
cat('<span style="color: #006400;">Removing', sum(is.na(coral_meow$realm)),
    'of', nrow(coral_meow), 'sample events due to missing realms</span>\n\n')

coral_meow <- coral_meow %>%
  filter(!is.na(realm)) #1191 sample events could not be connected to REALM
```

#### Get zonal statistics for SST (raster)

Raster covariates are extracted in two steps. [`get_zonal_statistics()`](https://data-mermaid.github.io/mermaidr-covariates/reference/get_zonal_statistics.html) returns one row per day per survey - here the 365 days before each sample date, spatially averaged within a 1 km radius (`spatial_stats = "mean"`) - and [`summarise_zonal_statistics()`](https://data-mermaid.github.io/mermaidr-covariates/reference/summarise_zonal_statistics.html) then collapses each survey's series to a single value. We take the **maximum**, so the variable is the hottest day of the year preceding each survey. Any window can be used via the `n_days` argument.

The chunk below runs the first step on 10 sample events so the daily structure is visible before it is collapsed; the full global extraction follows in the next chunk.

```{r get-zonal-stats-standard}
# Demonstrate the call on the first 10 sample events so the shape of the result
# is visible. The full extraction is done in the next chunk, in batches, and
# cached - so this example deliberately does not repeat that work.
# Look the covariate up by id rather than typing its title, which has changed
# upstream before.
sst_example <- coral_meow %>%
  slice(1:10) %>%
  mermaidrcovariates::get_zonal_statistics(
    covariate = "daily_sst",
    n_days = 365,
    radius = 1000,
    spatial_stats = "mean")

head(sst_example)
```

```{r get-zonal-stats-batches-checks-save}
#| results: asis
NumSe <- nrow(coral_meow)
NumDays <- 365
Radius <- 1000
batch_size <- 50  # Adjust based on what works for your data

# Define the file path for saved data. here::here() anchors this to the
# project's own data/ folder whether the code is run from a chunk, from the
# console, or during a render.
data_dir <- here::here("data")
if (!dir.exists(data_dir)) {
  dir.create(data_dir, recursive = TRUE)
}
sst_data_file <- file.path(data_dir, paste0("coralMeowSstSubset_NumSe", NumSe, "_NumDays", NumDays, ".rds"))

# Check if data already exists
if (file.exists(sst_data_file) & force_download == FALSE) {
  cat('<span style="color: #006400;">Loading previously downloaded SST data from:',
      sst_data_file, '</span>\n\n')
  coral_meow_sst <- readRDS(sst_data_file)
  cat('<span style="color: #006400;">Loaded',
      nrow(coral_meow_sst), 'SST records</span>\n\n')
} else {
  cat('<span style="color: #006400;">Downloading SST data (this may take a while)...',
      '</span>\n\n')

  # Start timer
  start_time <- Sys.time()

  # Process in batches to handle potential errors
  total_rows <- NumSe
  all_results <- list()
  failed_rows <- c()

  for (start in seq(1, total_rows, by = batch_size)) {
    end <- min(start + batch_size - 1, total_rows)

    cat('<span style="color: #006400;">Processing rows', start, 'to', end,
        'of', total_rows, '...</span>\n\n')

    tryCatch({
      # Summarise inside the batch. get_zonal_statistics() returns one row per
      # day per survey, so at 365 days across ~5,000 surveys the raw series is
      # ~1.8 million rows. Collapsing each batch first means all_results holds
      # one row per survey instead. The reduction is per survey, so this gives
      # exactly the same answer as summarising after binding.
      batch_result <- coral_meow %>%
        slice(start:end) %>%
        mermaidrcovariates::get_zonal_statistics(
          covariate = "daily_sst",
          n_days = NumDays,
          radius = Radius,
          spatial_stats = "mean"
        ) %>%
        summarise_zonal_statistics("max")
      all_results[[length(all_results) + 1]] <- batch_result

    }, error = function(e) {
      cat('<span style="color: #cc0000;">Batch error, processing row by row...</span>\n\n')

      # Process this batch row by row
      for (i in start:end) {
        tryCatch({
          row_result <- coral_meow %>%
            slice(i) %>%
            mermaidrcovariates::get_zonal_statistics(
              covariate = "daily_sst",
              n_days = NumDays,
              radius = Radius,
              spatial_stats = "mean"
            ) %>%
            summarise_zonal_statistics("max")
          all_results[[length(all_results) + 1]] <- row_result
        }, error = function(e2) {
          failed_rows <<- c(failed_rows, i)
          cat('<span style="color: #cc0000;">Row', i, 'failed</span>\n\n')
        })
      }
    })
  }

  # Combine all successful results
  coral_meow_sst <- bind_rows(all_results)

  # Calculate elapsed time
  end_time <- Sys.time()
  elapsed_time <- end_time - start_time

  # Print the result
  cat('<span style="color: #006400;">Data retrieval completed in:',
      round(elapsed_time, 2), attr(elapsed_time, "units"), '</span>\n\n')

  if (length(failed_rows) > 0) {
    cat('<span style="color: #cc0000;">Failed rows:',
        paste(failed_rows, collapse = ", "), '</span>\n\n')
  }

  # Unnest data for easier use. summarise_zonal_statistics() returns its
  # result in a `summary_zonal_statistics` list-column, as in the DHW example.
  coral_meow_sst <- coral_meow_sst %>%
    unnest(summary_zonal_statistics)

  cat('<span style="color: #006400;">Total SST records retrieved:',
      nrow(coral_meow_sst), '</span>\n\n')

  # Save the data
  saveRDS(coral_meow_sst, sst_data_file)
  cat('<span style="color: #006400;">Data saved to:',
      sst_data_file, '</span>\n\n')
}
```

## Create a final analysis file

```{r final-analysis-file}
#| results: asis
#Get rid of rows with NAs, select specific columns, and rename to more appropriate titles
analysis_data <- coral_meow_sst %>%
  # temporal_stats = "max" above means this is the highest daily SST in the
  # 365 days before each survey, spatially averaged within 1 km - a maximum,
  # not a mean, so name it accordingly.
  rename(sst_max = value) %>%
  select(project, country, site, sample_date,
         hard_coral_cover, sst_max,
         realm,
         latitude, longitude) %>%
  filter(!is.na(sst_max) & !is.na(hard_coral_cover) & !is.na(realm))

cat('<span style="color: #006400;">Number of SEs without NAs: ',
      nrow(analysis_data), '</span>\n\n')
```

## Data Summary

### Sample Distribution

```{r sample-summary}
# Summary by realm
realm_summary <- analysis_data %>%
  group_by(realm) %>%
  summarise(
    n_samples = n(),
    n_countries = n_distinct(country),
    n_sites = n_distinct(site),
    mean_coral_cover = mean(hard_coral_cover, na.rm = T),
    sd_coral_cover = sd(hard_coral_cover, na.rm = T),
    mean_max_sst = mean(sst_max, na.rm = T),
    sd_max_sst = sd(sst_max, na.rm = T),
    .groups = "drop"
  ) %>%
  arrange(desc(n_samples))

datatable(
  realm_summary,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Summary statistics by marine realm",
  rownames = FALSE
) %>%
  formatRound(columns = c("mean_coral_cover",
                          "sd_coral_cover",
                          "mean_max_sst", "sd_max_sst"), digits = 2)
```

## Coral Cover vs. SST Analysis

### Overall Relationship

```{r overall-relationship}
#| results: asis
# Overall scatter plot
overall_plot <- ggplot(analysis_data, aes(x = sst_max, y = hard_coral_cover)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "coral", linewidth = 1) +
  labs(
    x = "Maximum SST (°C)",
    y = "Hard Coral Cover (%)",
    title = "Hard Coral Cover vs. Sea Surface Temperature",
    subtitle = "Highest daily SST in the year before each survey, all marine realms"
  )

ggplotly(overall_plot)

# Calculate overall correlation
overall_cor <- cor.test(
  analysis_data$sst_max,
  analysis_data$hard_coral_cover
)

cat('<span style="color: #006400;">Overall correlation:', round(overall_cor$estimate, 3),
    '| P-value:', format.pval(overall_cor$p.value, digits = 3), '</span>\n\n')
```

### By Marine Realm

```{r by-realm}
# Filter to realms with sufficient sample sizes (e.g., n >= 20)
min_samples <- 20
realms_to_plot <- realm_summary %>%
  filter(n_samples >= min_samples) %>%
  pull(realm)

plot_data <- analysis_data %>%
  filter(realm %in% realms_to_plot)

# Create faceted scatter plot
realm_plot <- ggplot(plot_data, aes(x = sst_max, y = hard_coral_cover, color = realm)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  labs(
    x = "Maximum SST (°C)",
    y = "Hard Coral Cover (%)",
    title = "Hard Coral Cover vs. Sea Surface Temperature by Marine Realm",
    subtitle = paste("Realms with ≥", min_samples, "samples")
  ) +
  theme(legend.position = "bottom")

ggplotly(realm_plot)
```

### Realm-Specific Relationships

```{r realm-specific}
# Calculate correlations and linear model stats for each realm
realm_stats <- plot_data %>%
  group_by(realm) %>%
  summarise(
    n_samples = n(),
    correlation = cor(sst_max, hard_coral_cover, use = "complete.obs"),
    # Fit linear model
    slope = coef(lm(hard_coral_cover ~ sst_max))[2],
    intercept = coef(lm(hard_coral_cover ~ sst_max))[1],
    r_squared = summary(lm(hard_coral_cover ~ sst_max))$r.squared,
    p_value = summary(lm(hard_coral_cover ~ sst_max))$coefficients[2, 4],
    .groups = "drop"
  ) %>%
  arrange(desc(abs(correlation)))

datatable(
  realm_stats,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Linear relationship statistics by marine realm",
  rownames = FALSE
) %>%
  formatRound(columns = c("correlation", "slope", "intercept", "r_squared"), digits = 3) %>%
  formatSignif(columns = "p_value", digits = 3)
```

### Interactive Multi-Realm Plot

```{r interactive-realm-plot}
# Create interactive plot with all realms on same axes
interactive_plot <- plot_ly(data = plot_data) %>%
  add_markers(
    x = ~sst_max,
    y = ~hard_coral_cover,
    color = ~realm,
    colors = "Set1",
    text = ~paste(
      "Realm:", realm,
      "<br>Max SST:", round(sst_max, 2), "°C",
      "<br>Coral cover:", round(hard_coral_cover, 1), "%",
      "<br>Site:", site
    ),
    hoverinfo = "text",
    marker = list(size = 6, opacity = 0.5)
  ) %>%
  layout(
    title = "Hard Coral Cover vs. SST by Marine Realm",
    xaxis = list(title = "Maximum SST (°C)"),
    yaxis = list(title = "Hard Coral Cover (%)"),
    hovermode = "closest",
    legend = list(
      orientation = "v",
      x = 1.02,
      y = 1
    )
  )

# Add regression lines for each realm
for (realm_name in realms_to_plot) {
  realm_subset <- plot_data %>% filter(realm == realm_name)

  # Fit model
  model <- lm(hard_coral_cover ~ sst_max, data = realm_subset)

  # Create prediction data
  sst_range <- seq(min(realm_subset$sst_max), max(realm_subset$sst_max), length.out = 100)
  predictions <- predict(model, newdata = data.frame(sst_max = sst_range))

  # Add line
  interactive_plot <- interactive_plot %>%
    add_lines(
      x = sst_range,
      y = predictions,
      name = paste(realm_name, "trend"),
      line = list(dash = "dash"),
      showlegend = FALSE,
      hoverinfo = "skip"
    )
}

interactive_plot
```

## Key Findings

```{r summary-findings}
# Identify realms with strongest positive/negative relationships
strongest_positive <- realm_stats %>%
  filter(correlation > 0) %>%
  slice_max(correlation, n = 1)

strongest_negative <- realm_stats %>%
  filter(correlation < 0) %>%
  slice_min(correlation, n = 1)
```

**Overall Pattern:**

The global correlation between hard coral cover and maximum SST (the hottest day in the 365 prior to each survey) is `r round(overall_cor$estimate, 3)` (p = `r format.pval(overall_cor$p.value, digits = 3)`).

**Realm-Specific Patterns:**

- **Strongest positive relationship:** `r strongest_positive$realm` (r = `r round(strongest_positive$correlation, 3)`, n = `r strongest_positive$n_samples`)
- **Strongest negative relationship:** `r strongest_negative$realm` (r = `r round(strongest_negative$correlation, 3)`, n = `r strongest_negative$n_samples`)

The relationship between coral cover and SST varies considerably across marine realms, suggesting that local factors, thermal history, and biogeographic context play important roles in determining coral community composition and health.

## Data Access

This analysis uses:

- **MERMAID data:** Global benthic sample events (all protocols) accessed via the [`mermaidr`](https://github.com/data-mermaid/mermaidr) package
- **Environmental covariates:** Accessed via the [`mermaidrcovariates`](https://data-mermaid.github.io/mermaidr-covariates/) package
  - SST data: Daily Global 5km Satellite Sea Surface Temperature (CoralTemp), the maximum over the 365 days prior to each survey date, spatially averaged within a 1 km radius
  - Marine realms: Marine Ecoregions of the World (MEOW) classification, attached by location with `attach_covariate_data()`

## Session Info

```{r session-info}
sessionInfo()
```
