---
title: "Connecting MERMAID ecological data with covariates example"
subtitle: "Exploring relationships between hard coral cover and DHW in East Africa"
author:
  - name: Iain R. Caldwell
  - name: Sharla Gelfand
  - name: Emily Darling
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    self-contained: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
title-block-banner: "#f0f3f5"
title-block-banner-color: "black"
include-after-body: "footer.html"
editor_options: 
  chunk_output_type: console
---

## Overview

This analysis explores the relationship between hard coral cover and degree heating weeks (DHW) in East Africa. We use the [mermaidrcovariates](https://data-mermaid.github.io/mermaidr-covariates/) package to access environmental data for MERMAID survey locations.

**Key questions:**

- How does hard coral cover vary in East Africa?
- What is the exposure of different reefs to Degree Heating Weeks (DHW)?

## Setup

```{r setup}
#| warning: false
library(here)
library(tidyverse)
library(mermaidr)
library(plotly)
library(DT)
library(rstac)
library(geoarrow)
library(arrow)
library(sf)
library(janitor)

# Install mermaidr.covariates (if needed)
remotes::install_github("data-mermaid/mermaidr-covariates")

library(mermaidrcovariates)

# Set theme for ggplot
theme_set(theme_minimal(base_size = 12))

# Helper for the coloured status messages used throughout this document.
# knitr::asis_output() emits raw HTML on its own, so the chunk does NOT need
# `results: asis` and ordinary printed output (tibbles, data frames) in the
# same chunk still gets formatted normally.
status_msg <- function(..., color = "#006400") {
  knitr::asis_output(
    paste0('<span style="color: ', color, ';">', paste(...), '</span>\n\n')
  )
}

# Set this to TRUE to force re-download of MEOW boundaries
force_download <- FALSE
```

## Data Export

### MERMAID Benthic Data

Export all publicly available benthic PIT summary sample events from MERMAID.

```{r get-mermaid-data}
# Get all public benthic PIT data at the sample event level
mermaid_summ_ses_path <- here::here("data", "mermaid_summ_ses.rds")

if (file.exists(mermaid_summ_ses_path)) {
  mermaid_summ_ses <- readRDS(mermaid_summ_ses_path)
  status_msg("Loaded cached summary sample events from", mermaid_summ_ses_path,
             color = "#4682B4")
} else {
  mermaid_summ_ses <- mermaid_get_summary_sampleevents()
  dir.create(dirname(mermaid_summ_ses_path), showWarnings = FALSE, recursive = TRUE)
  saveRDS(mermaid_summ_ses, mermaid_summ_ses_path)
}

#file.exists("data/mermaid_summ_ses.rds")
#file.size("data/mermaid_summ_ses.rds")

mermaid_summ_ses <- mermaid_summ_ses %>% 
  filter(country %in% c("Kenya", "Tanzania"))

# Check how many sample events we have
status_msg("Total sample events:", nrow(mermaid_summ_ses))
```

### Extract Hard Coral Cover

Hard coral cover is stored in the percent cover columns. We'll focus on total hard coral cover and show a histogram of the data.

```{r extract-coral-cover}
# Extract relevant columns
coral_data <- mermaid_summ_ses %>% 
  rowwise() %>% 
  mutate(hard_coral_cover =
           mean(c(`benthicpit_percent_cover_benthic_category_avg_Hard coral`,
                  `benthiclit_percent_cover_benthic_category_avg_Hard coral`,
                  `benthicpqt_percent_cover_benthic_category_avg_Hard coral`,
                  quadrat_benthic_percent_percent_hard_avg_avg), na.rm = T)) %>% 
  ungroup() %>% 
  filter(!is.na(hard_coral_cover)) %>% 
  select(
    project,
    country,
    site,
    latitude,
    longitude,
    sample_date,
    hard_coral_cover
  ) %>% 
  # Convert date to Date format
  mutate(sample_date = as.Date(sample_date))

# Check how many sample events have hard coral cover estimates 
status_msg("Total sample events with hard coral data:", nrow(coral_data))

# Create the histogram to get bin data to assign colors
hist_data <- hist(coral_data$hard_coral_cover,
                  breaks = seq(0, 100, by = 2),
                  plot = FALSE) #In this function a 4 will be in the first bin

# Assign colors based on bin order
bin_colors <- sapply(seq_along(hist_data$counts), function(i) {
  if (i <= 5) {
    return('#d13823')
  } else if (i > 5 && i <= 15) {
    return('#f3a224')
  } else {
    return('#277d1d')
  }
})

# Create the histogram using plotly
hardCoralAggHist <-
  plot_ly(x = coral_data$hard_coral_cover,
          type = 'histogram',
          xbins = list(start = 0, size = 2, end = 100),
          marker = list(color = bin_colors), height = 450,
          hovertemplate = "Bin: %{x}%<br>%{y} surveys<extra></extra>") %>%
  config(displayModeBar = TRUE,
         displaylogo = FALSE,
         modeBarButtonsToRemove = c('zoom','pan', 'select', 'zoomIn', 'zoomOut',
                                    'autoScale', 'resetScale', 'lasso2d',
                                    'hoverClosestCartesian',
                                    'hoverCompareCartesian')) %>% 
  layout(bargap = 0.1,
         shapes = list(
           list(type = "line", 
                x0 = 10, x1 = 10, y0 = 0, y1 = 1, yref = "paper", 
                line = list(color = "black", dash = "dot")),
           list(type = "line", 
                x0 = 30, x1 = 30, y0 = 0, y1 = 1, yref = "paper", 
                line = list(color = "black", dash = "dot"))
         ),
         xaxis = list(title = "Hard coral cover (%)",
                      linecolor = "black",
                      linewidth = 2,
                      tickvals = seq(0, 100, by = 10),  # Set x-axis tick values
                      ticktext = seq(0, 100, by = 10)),
         yaxis = list(title = "Number of surveys",
                      linecolor = "black",   # Set the y-axis line color to black
                      linewidth = 2),
         annotations = list(
           list(x = 0, y = 1.15, text = "HARD CORAL COVER", showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 20)),
           list(x = 0, y = 1.08,
                text = paste0(nrow(coral_data), " Surveys"),
                showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 12))
         ),
         margin = list(t = 50, b = 75)) # Increase top margin to create more space for title and subtitle

# Visualize the plot
hardCoralAggHist
```

### Get Covariates

Now we'll use `mermaidrcovariates` to get DHW data.

#### List covariates

The first function of interest is [`list_covariates`](https://data-mermaid.github.io/mermaidr-covariates/reference/list_covariates.html), which can be used to find out which covariates are available.

```{r list-covariates}
# Extract id and title from list_covariates() output
covariates <- list_covariates()

covariates_df <- covariates %>% 
  select(id, title, description)

# Create interactive table
datatable(
  covariates_df,
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  caption = "Available MERMAID Covariates",
  rownames = FALSE,
  filter = "top"  # Adds search boxes at the top of each column
)
```

#### Get zonal statistics for DHW (raster)

The next function of interest is [get_zonal_statistics](https://data-mermaid.github.io/mermaidr-covariates/reference/get_zonal_statistics.html), which can be used to extract raster data, like DHW, from the covariates. In this case, we calculate the mean DHW for 30 days previous to the sample date, as an example, but it can be used to extract data across any number of days prior to the survey using the `n_days` argument in the function.

```{r get-zonal-stats-standard}

# How the surveys break down by country
coral_data %>% 
  tabyl(country) %>% 
  adorn_pct_formatting(digits = 1) %>% 
  rename(Country = country, Surveys = n, `Percent of surveys` = percent) %>% 
  knitr::kable(align = c("l", "r", "r"))

# get_zonal_statistics() needs the covariate's full title, so look it up by id
# rather than typing it out (pull() returns a plain string, not a tibble)
dhw_covariate <- covariates_df %>% 
  filter(id == "daily_dhw") %>% 
  pull(title)

status_msg("Using covariate:", dhw_covariate, color = "#4682B4")

#set path, use if/else to read or cache the API call
covars_path <- here::here("data", "coral_data_dhw_30days.rds")

# Start timer
start_time <- Sys.time()

if (file.exists(covars_path)) {
  
  coral_data_dhw <- readRDS(covars_path)
  status_msg("Loaded cached DHW data from", covars_path, color = "#4682B4")
  
} else {
  
  coral_data_dhw <- coral_data %>%
    mermaidrcovariates::get_zonal_statistics(
      covariate = dhw_covariate,
      n_days = 30,
      radius = 1000,
      spatial_stats = "mean")
  
  dir.create(dirname(covars_path), showWarnings = FALSE, recursive = TRUE)
  
  saveRDS(coral_data_dhw, covars_path)
  
}

# Calculate elapsed time
end_time <- Sys.time()
elapsed_time <- end_time - start_time

# Print the result
status_msg("Data retrieval completed in:",
           round(elapsed_time, 2), attr(elapsed_time, "units"))

```

```{r work-with-zonal-stats}
coral_data_dhw_summary <- coral_data_dhw %>%
  select(site, sample_date, zonal_statistics) %>%
  unnest(zonal_statistics)

#names(coral_data_dhw_summary)

# coral_data_dhw_summary %>%  
#   filter(value > 0) %>% 
#   arrange(-value)

max_dhw <- coral_data_dhw %>%
  summarise_zonal_statistics("max")

#max_dhw

max_dhw_summary <- max_dhw %>%
  select(site, sample_date, summary_zonal_statistics) %>%
  unnest(summary_zonal_statistics)

#max_dhw_summary
#names(max_dhw_summary)

dhw_4_or_more <- max_dhw_summary %>% 
  filter(value > 4) %>% 
  arrange(-value)

#nrow(dhw_4_or_more) #74 sites with DHW>4 in 30 days prior to survey

# Create the histogram to get bin data to assign colors
# Bin the data manually (single source of truth for bin edges)
bin_width <- 0.5
bin_start <- 0
bin_end <- ceiling(max(dhw_4_or_more$value))

breaks <- seq(bin_start, bin_end, by = bin_width)

dhw_binned <- dhw_4_or_more %>%
  mutate(bin = cut(value,
                   breaks = breaks,
                   right = FALSE,
                   include.lowest = TRUE)) %>%
  count(bin, .drop = FALSE) %>%
  mutate(
    bin_left = breaks[-length(breaks)],
    bin_mid  = bin_left + bin_width / 2,
    bin_color = if_else(bin_mid < 8, '#f3a224', '#d13823')
  )

# Create the histogram using plotly (as a bar chart on pre-binned data)
dhwAggHist <-
  plot_ly(data = dhw_binned, height = 450) %>%
  add_bars(x = ~bin_left + bin_width / 2,  # plot at bin midpoint
           y = ~n,
           width = bin_width * 0.9,        # bar width in x-axis units (trace attribute)
           marker = list(color = ~bin_color),
           hovertemplate = "Bin: %{x} DHW<br>%{y} surveys<extra></extra>") %>%
  config(displayModeBar = TRUE,
         displaylogo = FALSE,
         modeBarButtonsToRemove = c('zoom','pan', 'select', 'zoomIn', 'zoomOut',
                                    'autoScale', 'resetScale', 'lasso2d',
                                    'hoverClosestCartesian',
                                    'hoverCompareCartesian')) %>% 
  layout(bargap = 0.1,
         shapes = list(
           list(type = "line", 
                x0 = 8, x1 = 8, y0 = 0, y1 = 1, yref = "paper", 
                line = list(color = "black", dash = "dot"))
         ),
         xaxis = list(title = "Degree Heating Weeks (DHW)",
                      linecolor = "black",
                      linewidth = 2),
         yaxis = list(title = "Number of surveys",
                      linecolor = "black",
                      linewidth = 2),
         annotations = list(
           list(x = 0, y = 1.15, text = "DEGREE HEATING WEEKS", showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 20)),
           list(x = 0, y = 1.08,
                text = paste0(nrow(dhw_4_or_more), " Surveys"),
                showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 12))
         ),
         margin = list(t = 50, b = 75))

dhwAggHist
```

## Exposure to DHW by coral cover

```{r coral-cover-dhw}
#First make sure there are no duplicates in the max_dhw_summary data to join
max_dhw_summary_join <- max_dhw_summary |>
  filter(n_dates != 0) |> #get rid of any without data
  select(site, sample_date, covariate, value) |> 
  distinct() |> 
  group_by(site, sample_date, covariate) |> 
  summarise(value = mean(value)) |> 
  ungroup()

#Also remove duplicates in the coral data
coral_data_join <- coral_data |> 
  distinct() |> 
  group_by(project, country, site, latitude, longitude, sample_date) |> 
  summarise(hard_coral_cover = mean(hard_coral_cover)) |> 
  ungroup()

coral_data_with_covariate <- coral_data_join %>% 
  left_join(max_dhw_summary_join)

# Determine y-axis max to size the top band appropriately
y_max <- max(coral_data_with_covariate$value, na.rm = TRUE) * 1.05

# Create the scatter plot using plotly
coralDhwScatter <-
  plot_ly(data = coral_data_with_covariate,
          x = ~hard_coral_cover,
          y = ~value,
          type = 'scatter',
          mode = 'markers',
          marker = list(color = "black",
                        size = 8,
                        opacity = 0.6),
          height = 450,
          text = ~paste0(site, ", ", country, "<br>", format(sample_date, "%b %d, %Y")),
          hovertemplate = "%{text}<br>Hard coral cover: %{x:.1f}%<br>DHW: %{y:.1f}<extra></extra>") %>%
  config(displayModeBar = TRUE,
         displaylogo = FALSE,
         modeBarButtonsToRemove = c('zoom','pan', 'select', 'zoomIn', 'zoomOut',
                                    'autoScale', 'resetScale', 'lasso2d',
                                    'hoverClosestCartesian',
                                    'hoverCompareCartesian')) %>% 
  layout(shapes = list(
           # Background bands (drawn first, behind points)
           list(type = "rect",
                x0 = 0, x1 = 1, xref = "paper",
                y0 = 0, y1 = 4,
                fillcolor = '#277d1d', opacity = 0.12, line = list(width = 0),
                layer = "below"),
           list(type = "rect",
                x0 = 0, x1 = 1, xref = "paper",
                y0 = 4, y1 = 8,
                fillcolor = '#f3a224', opacity = 0.15, line = list(width = 0),
                layer = "below"),
           list(type = "rect",
                x0 = 0, x1 = 1, xref = "paper",
                y0 = 8, y1 = y_max,
                fillcolor = '#d13823', opacity = 0.15, line = list(width = 0),
                layer = "below"),
           # Threshold lines
           list(type = "line", 
                x0 = 0, x1 = 1, xref = "paper", y0 = 4, y1 = 4, 
                line = list(color = "black", dash = "dot")),
           list(type = "line", 
                x0 = 0, x1 = 1, xref = "paper", y0 = 8, y1 = 8, 
                line = list(color = "black", dash = "dot"))
         ),
         xaxis = list(title = "Hard coral cover (%)",
                      linecolor = "black",
                      linewidth = 2,
                      range = c(0, 100)),
         yaxis = list(title = "DHW exposure (past 30 days)",
                      linecolor = "black",
                      linewidth = 2,
                      range = c(0, y_max)),
         annotations = list(
           list(x = 0, y = 1.15, text = "CORAL COVER VS. HEAT STRESS", showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 20)),
           list(x = 0, y = 1.08,
                text = paste0(nrow(coral_data_with_covariate), " Surveys"),
                showarrow = FALSE, 
                xref = 'paper', yref = 'paper', xanchor = 'left', yanchor = 'top',
                font = list(size = 12))
         ),
         margin = list(t = 50, b = 75))

coralDhwScatter
```

## Facet by year - plot

```{r facet-by-year}
# Add year column (in case it wasn't already created upstream)
coral_data_with_covariate <- coral_data_with_covariate %>%
  mutate(year = lubridate::year(sample_date))

# Recreate y_max and bands so this chunk doesn't depend on earlier chunks
y_max <- max(coral_data_with_covariate$value, na.rm = TRUE) * 1.05

bands <- tibble(
  ymin = c(0, 4, 8),
  ymax = c(4, 8, y_max),
  fill = c('#277d1d', '#f3a224', '#d13823')
)

# Identify years where at least one survey had DHW > 4
years_with_heat_stress <- coral_data_with_covariate %>%
  filter(value > 4) %>%
  distinct(year) %>%
  pull(year)

# Filter to only those years (keeping all points within them)
coral_data_heat_years <- coral_data_with_covariate %>%
  filter(year %in% years_with_heat_stress)

coralDhwFacetGG <-
  ggplot(coral_data_heat_years, aes(x = hard_coral_cover, y = value)) +
  geom_rect(data = bands, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax, fill = fill),
            alpha = 0.15) +
  scale_fill_identity() +
  geom_hline(yintercept = c(4, 8), linetype = "dotted", color = "black") +
  geom_point(color = "black", alpha = 0.6, size = 1.8) +
  facet_wrap(~ year, ncol = 3) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, y_max)) +
  labs(title = "CORAL COVER VS. HEAT STRESS, BY YEAR",
       subtitle = paste0(nrow(coral_data_heat_years), " Surveys (years with DHW > 4 only)"),
       x = "Hard coral cover (%)",
       y = "DHW exposure (past 30 days)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey30"),
    axis.line = element_line(color = "black", linewidth = 0.6)
  )

coralDhwFacetGG
```

## Session Info

```{r session-info}
sessionInfo()
```
